A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-06-27 manuscript/manuscript.Rmd
RStudio freezing on me

@@

In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables. For dissolved oxygen (DO) and water temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, naive persistence and climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE. Yet, with chlorophyll-a, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE. Instead, BlockRNN and the naive persistence model attain the best performance with respect to CRPS and RMSE, respectively.

By examining some of the individual forecasts, as shown in Figure \ref{}, it is possible to gain intuition for why the neural network models perform well across the target variables. The AutoTheta model produces forecasts that resemble linear regressions based upon recently observed values. These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails catastrophically, negatively impacting the model's overall performance. For instance, after a peak of dissolved oxygen in the winter of 2024, the AutoTheta model forecasts that DO will continue to increase which is opposed to the trend that DO peaks in the winter and declines through the spring. Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in chlorophyll-a will lead to higher chlorophyll-a concentrations instead of an immediate reversion to the non-bloom state.

@@

These general patterns in model performance can be observed in the skill score plots displayed in Figures \ref{}. The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly. Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta as well as fewer forecasts that are outlier outperformers. The distributions of NN skill scores are more skewed towards outperforming than AutoTheta's skill scores are. In these plots, we display scores according to water body type, but we did not see any significant differences in performance across these categories. However, this may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.

When examining the performance of the models within a forecast horizon, additional nuances emerge, confusing the perspective that neural network models are the best performing model class. In Figure \ref{}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally. Meanwhile the neural network models underperform AutoTheta during the early stages of the forecast window (generally, when t < 5), but their forecast skill declines less rapidly than the skill score of AutoTheta. So while, the neural networks generally outperform AutoTheta and the null models according to coarse scoring metrics like mean CRPS and RMSE, AutoTheta is the best performer in short horizons.

@@

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_0.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_oxygen.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_oxygen.png}
  \end{subfigure}
  \caption{Skill score distributions for dissolved oxygen. The CRPSS plot measures the CRPS skill relative to the climatology model. The RMSE-SS plot measures the RMSE skill relative to the naive persistence model. A guideline is plotted at the threshold for underperformance: a score above this line denotes that a forecast is underperforming the reference model; a score below this line denotes that a forecast is outperforming the reference model. The AutoTheta model exhibits longer tails relative to most of the neural network models, indicating that AutoTheta has a larger amount of forecasts that do either very well or very poorly. The neural network models tend to have shorter tails in the underperforming region.}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_0.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_temperature.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_temperature.png}
  \end{subfigure}
  \caption{Skill score distributions for water temperature. AutoTheta does not have the long tails as seen in the skill score distributions for dissolved oxygen and chlorophyll-a, instead the neural network models have heavier tails in the outperforming region. This indicates that the neural network models are producing high accuracy forecasts more frequently than the AutoTheta model.}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_1.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_chla.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_chla.png}
  \end{subfigure}
  \caption{Skill score distributions for chlorophyll-a. The neural network models have shorter tails in the underperforming region relative to AutoTheta, indicating that the neural network models are producing a smaller amount of inaccurate forecasts. The naive persistence null model attained the lowest RMSE on chlorophyll-a, and this is apparent in the RMSE-SS plot as the models have skill score distributions that are generally centered above the underperformance threshold.}
\end{figure}

@@

  \caption{Mean skill scores within the 30-day forecast horizon. In the early phase of the forecast horizon (t < 5), the AutoTheta model is universally the top performing model. However, the performance of the AutoTheta model rapidly declines over the course of the 30-day horizon, concluding in AutoTheta being the worst performing model universally at the end of the horizon. The neural network models do not perform as well as AutoTheta at the beginning of the forecast horizon, but their performance declines less rapidly throughout the horizon.}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/forecast_legend.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_theta.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_theta.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_tft.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_tft.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_ensemble.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_ensemble.png}
  \end{subfigure}
  \caption{Individual forecasts of dissolved oxygen and chlorophyll-a. The AutoTheta model produces accurate forecasts most of the year, especially for dissolved oxygen; but, there are some intervals where AutoTheta produces extremely inaccurate forecasts. After peaks in dissolved oxygen and chlorophyll-a, the AutoTheta model wrongly predicts that the values will continue to increase. Conversely, the neural network-based forecasts from the TFT and the Naive Ensemble model have learned from the training data that high levels of dissolved oxygen in the winter are followed by a steady decline throughout the spring. With chlorophyll-a, the neural networks take a conservative strategy where they predict that chlorophyll-a will remain in a non-bloom state, and these models never take a risk in forecasting that a bloom will occur.}
\end{figure}

@@

The neural network models struggled to forecast chlorophyll-a, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{chen_combining}. With cholorophyll-a, the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state. It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation. Yet, it is also plausible that this conservative strategy was driven by the length of the forecast horizon. Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a. Predicting the timing of such an event in a long term horizon will be more difficult than performing the same prediction on a shorter horizon as there are more opportunities in the long term horizon to be wrong. With this reasoning, it is sensible that the neural networks would take a conservative approach.

In place of the one question that we set out to answer, namely, "How do neural network models perform for water quality forecasting?", we have unearthed an abundance of more questions and ideas for future work. Outside of exploring how the performance of neural networks will vary with forecast horizon, some other promising directions for exploration include cross-learning and hybrid models. Throughout this study, we have trained individual models for each target time series. Cross-learning presents a different approach where we would train one model across all the time series available for a given target variable. Through cross-learning, the model might be able to learn a pattern from one time series that will transfer to making accurate forecasts on a different time series. We also reckon that hybrid models may have superior forecasting performance to neural network or statistical models individually. We observed that AutoTheta were the best performing models early in the forecast horizon, and neural networks performed best later in the forecast horizon. This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.

>>>>

\begin{table}[!htbp]
    \centering
    \begin{subtable}{\textwidth}
        \centering
        \begin{tabular}{lcc}
            \toprule
            Abbreviation & Algorithm Name \\
            \midrule
            RNN & Recurrent Neural Network \cite{hochreiter_long_1997}  \\
            BlockRNN & Block Recurrent Neural Network \cite{du_multivariate_2020} \\
            TCN & Temporal Convolutional Network \cite{bai_empirical_2018} \\
            NLinear & NLinear \cite{zeng_are_2022}  \\
            DLinear &  DLinear \cite{zeng_are_2022} \\
            TFT & Temporal Fusion Transformer \cite{lim_temporal_2019} \\
            Transformer & Transformer \cite{vaswani_attention_2017} \\
        \end{tabular}
    \end{subtable}
    \caption{The neural network-based time series forecasting models used in this study.}
    \label{tab:ml_models}
\end{table}

@@

In Table \ref{tab:mean_crps_rmse}, we present the mean CRPS and RMSE scores for the respective target variables. For dissolved oxygen (DO) and water temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, naive persistence and climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE. Yet, with chlorophyll-a, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE. Instead, BlockRNN and the naive persistence model attain the best performance with respect to CRPS and RMSE, respectively.

By examining some of the individual forecasts, as shown in Figure \ref{fig:individual_forecasts}, it is possible to gain intuition for why the neural network models perform well across the target variables. The AutoTheta model produces forecasts that resemble linear regressions based upon recently observed values. These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails catastrophically, negatively impacting the model's overall performance. For instance, after a peak of dissolved oxygen in the winter of 2024, the AutoTheta model forecasts that DO will continue to increase which is opposed to the trend that DO peaks in the winter and declines through the spring. Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in chlorophyll-a will lead to higher chlorophyll-a concentrations instead of an immediate reversion to the non-bloom state.

@@

These general patterns in model performance can be observed in the skill score plots displayed in Figures \ref{fig:oxygen_skill}, \ref{fig:temperature_skill}, \ref{fig:chla_skill}. The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly. Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta as well as fewer forecasts that are outlier outperformers. The distributions of NN skill scores are more skewed towards outperforming than AutoTheta's skill scores are. In these plots, we display scores according to water body type, but we did not see any significant differences in performance across these categories. However, this may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.

When examining the performance of the models within a forecast horizon, additional nuances emerge, confusing the perspective that neural network models are the best performing model class. In Figure \ref{fig:intra_skill}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally. Meanwhile the neural network models underperform AutoTheta during the early stages of the forecast window (generally, when t < 5), but their forecast skill declines less rapidly than the skill score of AutoTheta. So while, the neural networks generally outperform AutoTheta and the null models according to coarse scoring metrics like mean CRPS and RMSE, AutoTheta is the best performer in short horizons.

@@

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_0.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_oxygen.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_oxygen.png}
  \end{subfigure}
  \caption{Skill score distributions for dissolved oxygen. The CRPSS plot measures the CRPS skill relative to the climatology model. The RMSE-SS plot measures the RMSE skill relative to the naive persistence model. A guideline is plotted at the threshold for underperformance: a score above this line denotes that a forecast is underperforming the reference model; a score below this line denotes that a forecast is outperforming the reference model. The AutoTheta model exhibits longer tails relative to most of the neural network models, indicating that AutoTheta has a larger amount of forecasts that do either very well or very poorly. The neural network models tend to have shorter tails in the underperforming region.}
  \label{fig:oxygen_skill}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_0.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_temperature.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_temperature.png}
  \end{subfigure}
  \caption{Skill score distributions for water temperature. AutoTheta does not have the long tails as seen in the skill score distributions for dissolved oxygen and chlorophyll-a, instead the neural network models have heavier tails in the outperforming region. This indicates that the neural network models are producing high accuracy forecasts more frequently than the AutoTheta model.}
  \label{fig:temperature_skill}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/sitetype_legend_1.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_historical_chla.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/type_local_naive_chla.png}
  \end{subfigure}
  \caption{Skill score distributions for chlorophyll-a. The neural network models have shorter tails in the underperforming region relative to AutoTheta, indicating that the neural network models are producing a smaller amount of inaccurate forecasts. The naive persistence null model attained the lowest RMSE on chlorophyll-a, and this is apparent in the RMSE-SS plot as the models have skill score distributions that are generally centered above the underperformance threshold.}
  \label{fig:chla_skill}
\end{figure}

@@

  \caption{Mean skill scores within the 30-day forecast horizon. In the early phase of the forecast horizon (t < 5), the AutoTheta model is universally the top performing model. However, the performance of the AutoTheta model rapidly declines over the course of the 30-day horizon, concluding in AutoTheta being the worst performing model universally at the end of the horizon. The neural network models do not perform as well as AutoTheta at the beginning of the forecast horizon, but their performance declines less rapidly throughout the horizon.}
  \label{fig:intra_skill}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/forecast_legend.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_theta.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_theta.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_tft.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_tft.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/oxygen_ensemble.png}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{./plots/chla_ensemble.png}
  \end{subfigure}
  \caption{Individual forecasts of dissolved oxygen and chlorophyll-a. The AutoTheta model produces accurate forecasts most of the year, especially for dissolved oxygen; but, there are some intervals where AutoTheta produces extremely inaccurate forecasts. After peaks in dissolved oxygen and chlorophyll-a, the AutoTheta model wrongly predicts that the values will continue to increase. Conversely, the neural network-based forecasts from the TFT and the Naive Ensemble model have learned from the training data that high levels of dissolved oxygen in the winter are followed by a steady decline throughout the spring. With chlorophyll-a, the neural networks take a conservative strategy where they predict that chlorophyll-a will remain in a non-bloom state, and these models never take a risk in forecasting that a bloom will occur.}
  \label{fig:individual_forecasts}
\end{figure}

@@

The neural network models struggled to forecast chlorophyll-a, a result that is not surprising given that forecasting algae blooms is well established as a challenging prediction problem in limnology \cite{chen_combining}. With chlorophyll-a, the neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chla concentration would remain in a non-bloom state. It is possible that the neural network models settled on this conservative behavior because we did provide enough information to predict blooms; the neural networks were not provided with some of the typical covariates used in process-based models like Nitrogen and Phosphorous concentrations, and photosynthetic active radiation. Yet, it is also plausible that this conservative strategy was driven by the length of the forecast horizon. Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a. Predicting the timing of such an event in a long term horizon will be more difficult than performing the same prediction on a shorter horizon as there are more opportunities in the long term horizon to be wrong. With this reasoning, it is sensible that the neural networks would take a conservative approach.

This study had the simple motivation of comducting a model comparison, and has resulted in an abundance of research questions and ideas for future work. Outside of exploring how the performance of neural networks will vary with forecast horizon, other promising directions for exploration include cross-learning and hybrid models. Throughout this study, we trained individual models for each target time series. Cross-learning presents a different approach where we would train one model across multiple time series (potentially, all the time series for a give target variable or even across target variables). Through cross-learning, the model might be able to learn patterns from one time series that will transfer to making accurate forecasts on a different time series. Along these lines, developing a model hybridized across modeling classes could yield superior performance. Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model alone. For instance, one could explore a model that heavily relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the  when  longer horizons. This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.
